---
title: "Missing Value Imputation"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

-----


##### Load the data set

```{r Load}
INFILE <- "C:/Users/aroon/Desktop/MSDS/422/MSDS422/Week 1/HMEQ_Loss.csv"
df <- read.csv(INFILE)
copy.df <- df

```
##### Preliminary exploration of dataset

```{r exploration}

# Exploring Raw data
head(df) # first 6
tail(df) # last 6

# view the class of the data input
class(df)
print(str(df)) # structure of data frame

# view diamension #rows and columns
dim(df)
nrow(df)
names(df) # column names
summary(df) # summary of Data Frame
sum(complete.cases(df)) # Check for number of complete rows
summary(df$LOAN)
summary(df$MORTDUE)
# load dplyr
library(dplyr)
glimpse(df)
# find the mean of debt to income ratio by variable BAD
print( with( df, tapply( DEBTINC, TARGET_BAD_FLAG, mean, na.rm=TRUE ) ) )
# find the mean of mortgage due by variable BAD
print( with( df, tapply( MORTDUE, TARGET_BAD_FLAG, mean, na.rm=TRUE ) ) )

```



##### Graphs

```{r Graphs}
par(mfrow = c(3,3))

hist(df$TARGET_BAD_FLAG, xlab="Bad loans", main=" BAD loans") # histogram of bad loans
#Amount of the loan request
hist(df$LOAN,main="Loans")
plot(df$MORTDUE,df$LOAN)
hist(df$VALUE/df$MORTDUE ,main= "Value to Mortgage Ratio")
# find the mean of debt to income ratio by variable BAD
barplot( with( df, tapply( DEBTINC, TARGET_BAD_FLAG, mean, na.rm=TRUE ,main="Debt to income classified by Bad debt") ) )






df$M_DEBTINC = is.na( df$DEBTINC ) + 0
df$M_VALUE = is.na( df$VALUE ) + 0
df$M_MORTDUE = is.na( df$MORTDUE ) + 0
df$M_YOJ = is.na( df$YOJ ) + 0
df$M_DEROG = is.na( df$DEROG ) + 0
df$M_DELINQ = is.na( df$DELINQ ) + 0
df$M_CLAGE = is.na( df$CLAGE ) + 0
df$M_NINQ = is.na( df$NINQ ) + 0
df$M_CLNO = is.na( df$CLNO ) + 0

head(df)

```
##### Find the mean of the columns. Copy the column value into new column and call it IMP that will be corrected values. 


```{r Mean_COPYCOLUMN data}
#mean( df$DEBTINC, na.rm=TRUE ) # used to find mean.
df$IMP_DEBTINC = df$DEBTINC
df$IMP_VALUE = df$VALUE
df$IMP_MORTDUE = df$MORTDUE
df$IMP_YOJ = df$YOJ
df$IMP_DEROG = df$DEROG
df$IMP_DELINQ = df$DELINQ
df$IMP_CLAGE =  df$CLAGE
df$IMP_NINQ =  df$NINQ
df$IMP_CLNO = df$CLNO

```
##### Create new column for every column that will have imputed values and have column name with IMP prefixed. Fill the missing values with the average value.


```{r column IMP_}

####impute using mean
df$IMP_DEBTINC = ifelse(is.na( df$IMP_DEBTINC ), mean( df$DEBTINC, na.rm=TRUE ), df$IMP_DEBTINC ) 
df$IMP_CLAGE = ifelse(is.na(df$CLAGE), mean( df$CLAGE, na.rm= TRUE ), df$IMP_CLAGE )   
df$IMP_YOJ = ifelse(is.na(df$YOJ), mean( df$YOJ, na.rm= TRUE ), df$IMP_YOJ )
# though there is only 1 maximum value of 10 and mean = 0.2546 median is 0. It's safe to assume the mean value for the missing value.
df$IMP_DEROG = ifelse(is.na(df$DEROG), mean( df$DEROG, na.rm= TRUE ), df$IMP_DEROG )
# safer to assume mean value when the data is missing because median, 1st,3rd quartile value are 0
df$IMP_DELINQ = ifelse(is.na(df$DELINQ), mean( df$DELINQ, na.rm= TRUE ), df$IMP_DELINQ ) 

####impute using median 

df$IMP_VALUE = ifelse(is.na( df$IMP_VALUE ), median( df$VALUE, na.rm=TRUE ), df$IMP_VALUE )
df$IMP_MORTDUE = ifelse(is.na( df$MORTDUE), median( df$MORTDUE, na.rm=TRUE),df$IMP_MORTDUE ) 
# Number of recent credit inquiries are usually whole numbers. mean is only.1 more than median.
df$IMP_NINQ = ifelse(is.na(df$NINQ), median( df$NINQ, na.rm= TRUE ), df$IMP_NINQ ) 
# number of credit line is a whole number and since mean is only .1 greater than median.I used median value
df$IMP_CLNO = ifelse(is.na(df$CLNO), mean( df$CLNO, na.rm= TRUE ), df$IMP_CLNO ) 

```
##### Getting rid of the columns with missing values
```{r subset}
print( head(df) )
df = subset(df, select = -c( DEBTINC,VALUE,MORTDUE,YOJ,CLAGE,DEROG,DELINQ,NINQ,CLNO ) )
print(head(df))

```
##### Handle Categorical variable
```{r categorical}
print( with( df, tapply( TARGET_BAD_FLAG, JOB, mean, na.rm=TRUE ) ) ) # get the bad loan based on job title.
df$IMP_JOB = df$JOB # copy JOB into IMP_JOB
df$IMP_REASON = df$REASON # copy REASON into IMP_REASON
# id the job is blank then mark it unknown

df$IMP_JOB = ifelse(df$JOB == "", "UNKNOWN", as.character(df$IMP_JOB) )
df$IMP_REASON = ifelse(df$REASON == "", "UNKNOWN", as.character(df$IMP_REASON) )

print( head(df) )
df = subset(df, select = -c( JOB, REASON ) )
print( head(df) )


```
##### * Outlier Stats
```{r Outlier}
boxplot(copy.df)
par(mfrow = c(1,2))
# the 2 variables with outliers that needs to be fixed 
boxplot(df$IMP_DEBTINC,main="Debt to income ratio") 
boxplot(df$IMP_MORTDUE, main="Mortgage due") 


```
##### Get the MAX, MIN, MEAN, SD for 2 variables - DEBTINC and MARTDUE
```{r boxplotstats}
a1 = max( df$IMP_DEBTINC, na.rm=TRUE )
z1 = min( df$IMP_DEBTINC, na.rm=TRUE )
m1 = mean( df$IMP_DEBTINC, na.rm=TRUE )
s1 = sd( df$IMP_DEBTINC, na.rm=TRUE )
a2 = max( df$IMP_MORTDUE, na.rm=TRUE )
z2 = min( df$IMP_MORTDUE, na.rm=TRUE )
m2 = mean( df$IMP_MORTDUE, na.rm=TRUE )
s2 = sd( df$IMP_MORTDUE, na.rm=TRUE )

```
##### If the value beyond  mean +3 standard deviation then set it to mean+3*SD and if value is lesser than mean-3SD ,then replace it with mean-3SD.
```{r replaceOutlier}
df$IMP_DEBTINC = ifelse( df$IMP_DEBTINC > m1+3*s1, m1+3*s1, df$IMP_DEBTINC )
df$IMP_DEBTINC = ifelse( df$IMP_DEBTINC < m1-3*s1, m1-3*s1, df$IMP_DEBTINC )
df$IMP_MORTDUE = ifelse( df$IMP_MORTDUE > m2+3*s2, m2+3*s2, df$IMP_MORTDUE )
df$IMP_MORTDUE = ifelse( df$IMP_MORTDUE < m2-3*s2, m2-3*s2, df$IMP_MORTDUE )

```
##### Histogram and box plot after fixing outliers
```{r graph_afterOutlier}
par(mfrow=c(2,3))
hist( df$IMP_DEBTINC  )
hist( df$IMP_MORTDUE  )
boxplot( df$IMP_DEBTINC ,main= "Debt to income ratio" )
boxplot( df$IMP_MORTDUE , main="Mortgage due"  )

```
##### Recalculate the MAX, MIN, MEAN, SD for 2 variables - DEBTINC and MARTDUE
```{r recal_values}
a1 = max( df$IMP_DEBTINC, na.rm=TRUE )
z1 = min( df$IMP_DEBTINC, na.rm=TRUE )
m1 = mean( df$IMP_DEBTINC, na.rm=TRUE )
s1 = sd( df$IMP_DEBTINC, na.rm=TRUE )
a2 = max( df$IMP_MORTDUE, na.rm=TRUE )
z2 = min( df$IMP_MORTDUE, na.rm=TRUE )
m2 = mean( df$IMP_MORTDUE, na.rm=TRUE )
s2 = sd( df$IMP_MORTDUE, na.rm=TRUE )
```
##### Create table and pie chart of the categorical table
```{r categorical_graph_}
tbl1 <- table( df$IMP_JOB )
tbl2 <- table( df$IMP_REASON )
par(mfrow=c(1,2))
pie( tbl1 , main= "JOB" )
pie( tbl2, main= "REASON" )
```



##### Create new variable to identift risky jobs
```{r newColumn_}
print( with( df, tapply( TARGET_BAD_FLAG, IMP_JOB, mean ) ) )
df$FLAG_JOB_RISKY = ifelse(df$IMP_JOB %in% c("Self","Sales"), 1, 0 )
head(df,n=25)
summary(df)
```
##### Prediction using decision tree
```{r DecisionTree_}

library( rpart ) # model
library( rpart.plot )
df2 <- df 
t = rpart( TARGET_BAD_FLAG ~. , method="class", data=df)
rpart.plot( t, box.palette="RdBu", shadow.col="gray", nn=TRUE) 
p_bad = predict( t, data=df2, type=c("class") )
p_Prob = predict( t, data=df2, type=c("prob") )
df2 = cbind( df2, p_bad )
df2 = cbind( df2, p_Prob )

accuracy = sum(df2$BAD == df2$p_bad)/length(df2$BAD)
cat( "The accuracy of prediction is",accuracy )
```